Evaluating FAIR Digital Object as a distributed object system

### 2023-02-01 latex/main.tex
more references sorted

% https://www.overleaf.com/latex/examples/bibliographies-with-biber-and-biblatex/ccrkczqwnywf
% use biblatex/biber as natbib is from 1990s and have no DOIs
\usepackage[autostyle]{csquotes}
\usepackage[backend=biber,style=authoryear]{biblatex}
\addbibresource{bibliography.bib}
\DeclareFieldFormat{doi}{%
  \url{https://doi.org/#1}}

@@

\author[1,2]{Stian Soiland-Reyes}
\author[1]{Carole Goble}
\author[2]{Paul Groth}
\affil[1]{Department of Computer Science, The University of Manchester, UK}
\affil[2]{Informatics Institute, Faculty of Science, University of Amsterdam, NL }
\corrauthor[1]{Stian Soiland-Reyes}{soiland-reyes@manchester.ac.uk}

@@

The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee \cite{Guy7cjVY}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs. The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}. The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.

Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \cite{16WGmCih8} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \cite{16wvFH4nh} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).

While triple-based knowledge representations were not new \cite{NxvhwSmG}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \cite{RRtwPF7Y} are generalised forms of URLs that include locator-less identifiers such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years \cite{nW4HY8Nq}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \cite{HqZ3r6J3}. RDF 1.1 formats use Unicode to support IRIs \cite{rqjuXkEf}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \cite{rnzJ4khD}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.

@@

The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is). More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \cite{yd2WE9eq} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.

Established Semantic Web services and standards include SPARQL \cite{xww8fb1B} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).

The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins \cite{bMSxa8XW,Qlrlr2TZ}. In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same. One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.

Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies. Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS \cite{15gQDya5B} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties). Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \cite{FDDmgO6s} .

The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources. This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \cite{191ZkYZMt}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \cite{9CdJnsoJ}. Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \cite{AGHR9oyr}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \cite{5SuwLNdE}. Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \cite{ZwAcGQKY}, OWL \cite{1p4IWJpI}, OBO \cite{1DalZuKct}, SKOS \cite{15gQDya5B}), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.

@@

%\bibliography{bibliography}
\printbibliography

>>>>

% Make tt font not so massive
\renewcommand*\ttdefault{cmvtt}

% https://www.overleaf.com/latex/examples/bibliographies-with-biber-and-biblatex/ccrkczqwnywf
% use biblatex/biber as natbib is from 1990s and have no DOIs
\usepackage[autostyle]{csquotes}
\usepackage[backend=biber,style=authoryear]{biblatex}
\addbibresource{bibliography.bib}
\DeclareFieldFormat{doi}{%
  \url{https://doi.org/#1}}
%\urlstyle{tt}

@@

\author[1,2]{Stian Soiland-Reyes} % https://orcid.org/0000-0001-9842-9718
\author[1]{Carole Goble}          % https://orcid.org/0000-0003-1219-2137
\author[2]{Paul Groth}            % https://orcid.org/0000-0003-0183-6910

\affil[1]{Department of Computer Science, The University of Manchester, UK}
\affil[2]{Informatics Institute, Faculty of Science, University of Amsterdam, NL }
\corrauthor[1]{Stian Soiland-Reyes}{soiland-reyes@manchester.ac.uk}

@@

The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee \autocite{berners-leeWeavingWebOriginal1999}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs. The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}. The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.

Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \autocite{w3-rdf11-primer} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \autocite{w3-rdf-syntax} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).

While triple-based knowledge representations were not new \autocite{stanczykProcessModellingInformation1987}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \autocite{rfc3986} are generalised forms of URLs that include locator-less identifiers such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years \autocite{InfoURIRegistry}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \autocite{jutyIdentifiersOrgMIRIAM2011}. RDF 1.1 formats use Unicode to support IRIs \autocite{rfc3987}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \autocite{sauermannCoolURIsSemantic2011}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.

@@

The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is). More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \autocite{horrocksSemanticWebISWC2002} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.

Established Semantic Web services and standards include SPARQL \autocite{w3-sparql11-overview} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).

The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins \autocite{gobleStateNationData2008c,williamsOpenPHACTSSemantic2012c}. In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same. One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.

Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies. Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS \autocite{w3-skos-primer} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties). Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \autocite{huHowMatchableAre2011a} .

The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources. This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \autocite{grothAPIcentricLinkedData2014b}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \autocite{kamdarSystematicAnalysisTerm2017a}. Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \autocite{kleinScholarlyContextNot2014a}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \autocite{polleresMoreDecentralizedVision2020a}. Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \autocite{carrieroLandscapeOntologyReuse2020a}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \autocite{w3-rdf-schema}, OWL \autocite{w3-owl2-overview}, OBO \autocite{tirmiziMappingOBOOWL2011a}, SKOS \autocite{w3-skos-primer}), hash vs slash, HTTP status codes and PID redirection strategies \autocite{sauermannCoolURIsSemantic2011}.

@@

%\bibliography{bibliography}
\def\UrlFont{\small}
\printbibliography
